### 2011-06-06 docs/pubs/EIM-2011/application.tex
further refinements: -use WET mass attribute in text since it appears in the figures/screenshots -finesse query example section, break long sentences into two -resolve DML citation link (was broken)

As discussed above, by leveraging the relationships defined and
inferred from the ontology we are able to increase recall beyond what
is possible for simple keyword-based searches
\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
direct matches as well as subclass matches. The queries can be quickly
refined when using the Web application by allowing rapid exploration
of the data repository without having to define complete observational
queries \emph{de novo}. The interface allows users to specify
individual classes of a measurement as well as pre-configured
measurement types (representing standard data-set attribute types) as
defined in OBOE compatible ontologies to enable a single concept to
proxy its constituent parts, namely the characteristics of particular
entities that can be measured with a set of protocols and
standards. This short-hand query generation can save users time in
specifying their queries, and highlights a compelling reason for using
OBOE extension ontologies and can also be leveraged when creating or
editing semantic annotations in the Morpho interface.

@@

\mypara{Data-level query.} For even more precise recall, the OBOE
model can be (partially) materialized (see above) during the query
stage after which a data range filter can be applied. Different
techniques are available for merging the annotation with the data that
it describes, but for performance reasons a hybrid approach has been
adopted in which preliminary search results from a structured query
are collated and only that subset is materialized. Because our corpus
is described using EML in conjunction with the annotation syntax, the
Data Manager Library \cite{leinfelder10} is used to load the described
raw data (into a relational database) while the annotation informs the
correct use of the Data Manager query and filtering features. For any
measurements that match the concept query criteria, we verify that
those measurements (e.g., attributes) contain data values within the
range specified in the initial semantic data query and return the data
packages that contain them (see \figref{fig:metacat-query}).

@@

\begin{figure}
\centering
\includegraphics[scale=1.0]{images/integration.png}
\caption{Data integration query across multiple data packages (A, B).
  Annotations (C, D) are used to determine semantically equivalent
  data attributes contained in the data objects (E, F). Attributes
  'plot' and 'site' are considered equivalent measurements of the
  characteristic 'Location'; attributes 'weight' and 'wt' both map to
  the same characteristic 'Mass'. The Semantic Mediation API utilizes
  the Data Manager Library to load and query the source data informed
  by semantic similarities between the structurally disparate data
  attributes.}
\label{fig:integration}
\end{figure}

>>>>

As discussed above, by leveraging the relationships defined and
inferred from the ontology we are able to increase recall beyond what
is possible for simple keyword-based searches
\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
direct matches as well as subclass matches. The queries can be quickly
refined when using the Web application by allowing rapid exploration
of the data repository without having to define complete observational
queries \emph{de novo}. The interface allows users to specify
individual classes of a measurement as well as pre-configured
measurement types (representing standard data-set attribute types) as
defined in OBOE compatible ontologies to enable a single concept to
proxy its constituent parts, namely the characteristics of particular
entities that can be measured with a set of protocols and
standards. This short-hand query generation can save users time in
specifying their queries, and highlights a compelling reason for using
OBOE extension ontologies. Measurement templates can also be leveraged when creating or
editing semantic annotations in the Morpho interface.

@@

\mypara{Data-level query.} For even more precise recall, the OBOE
model can be (partially) materialized (see above) during the query
stage after which a data range filter can be applied. Different
techniques are available for merging the annotation with the data that
it describes, but for performance reasons a hybrid approach has been
adopted in which preliminary search results from a structured query
are collated and only that subset is materialized. Because our corpus
is described using EML in conjunction with the annotation syntax, the
Data Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin} is used to load the described
raw data (into a relational database) while the annotation informs the
correct use of the Data Manager query and filtering features. For any
measurements that match the concept query criteria, we verify that
those measurements (e.g., attributes) contain data values within the
range specified in the initial semantic data query and return the data
packages that contain them (see \figref{fig:metacat-query}).

@@

\begin{figure}
\centering
\includegraphics[scale=1.0]{images/integration.png}
\caption{Data integration query across multiple data packages (A, B).
  Annotations (C, D) are used to determine semantically equivalent
  data attributes contained in the data objects (E, F). Attributes
  {\tt plot} and {\tt site} are considered equivalent measurements of the
  characteristic Location; attributes {\tt weight} and {\tt wt} both map to
  the same characteristic Mass. The Semantic Mediation API utilizes
  the Data Manager Library to load and query the source data informed
  by semantic similarities between the structurally disparate data
  attributes.}
\label{fig:integration}
\end{figure}

### 2011-06-06 docs/pubs/EIM-2011/framework.tex
further refinements: -use WET mass attribute in text since it appears in the figures/screenshots -finesse query example section, break long sentences into two -resolve DML citation link (was broken)

The Semtools project has focused development efforts on three main
components: a Java library for accessing and manipulating OBOE
ontology extensions and semantic annotations, an annotation plugin for
the Morpho metadata editor, and query extensions for the Metacat data
catalog. Below we briefly describe these components and give a brief
overview of the OBOE model as well as the semantic annotation approach
used in Semtools. For a more in-depth presentation of OBOE see
\cite{madin07:_ontol_for_descr_and_synth,bowers08}.

@@

Semantic annotations can be used to facilitate discovery and
integration of heterogeneous data sets. For instance, combining
semantic annotations with OBOE, it is possible to discover data sets
based on searches expressed over types of observations and
measurements of interest. As simple examples, users can pose queries
such as ``\emph{find all data sets containing observations of Kelp
  entities}'' and ``\emph{find all data sets containing Mass
  measurements of Kelp}''.\footnote{Note that queries also can include
  contextual constraints, e.g., Mass measurements of Kelp within a
  specific spatial or temporal context.} Both of these queries would
return the example data set in \figref{fig:kelp-mass-model} since the
attribute {\tt DRY} is linked to a DryMass measurement of observations
of Macrocystis (where DryMass is defined as a special kind, or
\emph{subclass} of Mass, and Macrocystis is defined as a subclass of
Kelp). Using semantic annotations in this way can help to increase
both query recall and precision over standard keyword-based approaches
\cite{berkley09:_improv_data_discov_for_metad}. In particular, by
defining terms as subclasses of other terms (e.g., Macrocystis as a
subclass of Kelp), term expansion can be used to increase the number
(recall) of data sets returned (where subclasses of query terms are
also searched), and the precision of the result can be improved since
queries can specify the desired connections between terms (e.g.,
measurements of Mass for Kelp observations) as opposed to returning
all data sets that simply mention the terms but without any explicit
connections (i.e., where Mass was measured, but not for Kelp samples).
Annotations also help facilitate integration by allowing tools to
align data-set attributes based on their declared measurement and
observation types.

In general, semantic annotations provide a formal description of
attribute semantics, whereas in many commonly-used metadata formats
for describing data sets, only informal text-based descriptions of
data attributes are permitted.  In the case of EML, there is some
overlap between these two mechanisms---particularly with respect to
measurement standards---however, semantic annotations extend this
approach by providing a general mechanism to formally associate
concepts drawn from domain ontologies to attributes. In the Semtools
framework, we employ an XML serialization syntax for semantic
annotations that is compatible with EML but that is stored separately
from the EML documentation of a data set (allowing, e.g., annotations
to be used independently of EML or with other metadata standards if
needed). In addition, semantic annotations can be used to
``materialize'' a given data set into a set of triples conforming to
the model configuration given in the annotation. In other words, a
tabular data set such as the one shown in \figref{fig:kelp-mass-model}
can automatically be converted into a corresponding collection of
observation and measurement instances. This in turn enables a simple
form of structural integration, where instead of having a large number
of different tabular data structures, all data is represented using
the standard set of structures defined by the OBOE model (see
\figref{fig:oboe}). Thus, materializing a data set in this way
provides a more uniform structural representation that can make a
number of discovery and integration tasks easier. For instance,
materialization can be used to increase query expressivity by allowing
searches of the form ``\emph{find all data sets containing Mass
  measurements of Kelp with values less than 1 Kilogram}'', which in
our example can be answered by generating (i.e., materializing) the
measurements associated with the {\tt DRY} attribute in the data set
of \figref{fig:kelp-mass-model}.

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows users to define annotations for existing data package
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierarchical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data
packages with semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated and fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data-set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., observations). The editor
provides a number of additional features including allowing users to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements.

>>>>

The Semtools project has focused development efforts on three main
components: a Java library for accessing and manipulating OBOE
ontology extensions and semantic annotations, an annotation plugin for
the Morpho metadata editor, and query extensions for the Metacat data
catalog. Below we describe these components and give a brief
overview of the OBOE model as well as the semantic annotation approach
used in Semtools. For a more in-depth presentation of OBOE see
\cite{madin07:_ontol_for_descr_and_synth,bowers08}.

@@

Semantic annotations can be used to facilitate discovery and
integration of heterogeneous data sets. For instance, combining
semantic annotations with OBOE, it is possible to discover data sets
based on searches expressed over types of observations and
measurements of interest. As simple examples, users can pose queries
such as ``\emph{find all data sets containing observations of Kelp
  entities}'' and ``\emph{find all data sets containing Mass
  measurements of Kelp}''.\footnote{Note that queries also can include
  contextual constraints, e.g., Mass measurements of Kelp within a
  specific spatial or temporal context.} Both of these queries would
return the example data set in \figref{fig:kelp-mass-model} since the
attribute {\tt WET} is linked to a WetMass measurement for observations
of Macrocystis (where WetMass is defined as a special kind, or
\emph{subclass} of Mass, and Macrocystis is defined as a subclass of
Kelp). Using semantic annotations in this way can help to increase
both query recall and precision over standard keyword-based approaches
\cite{berkley09:_improv_data_discov_for_metad}. In particular, by
defining terms as subclasses of other terms (e.g., Macrocystis as a
subclass of Kelp), term expansion can be used to increase the number
(recall) of data sets returned (where subclasses of query terms are
also searched). The precision of the result can be improved since
queries may specify the desired connections between terms (e.g.,
measurements of Mass for Kelp observations) as opposed to returning
all data sets that simply mention the terms but without any explicit
connections (i.e., where Mass was measured, but not for Kelp samples).
Annotations also help facilitate integration by allowing tools to
align data-set attributes based on their declared measurement and
observation types.

In general, semantic annotations provide a formal description of
attribute semantics, whereas in many commonly-used metadata formats
for describing data sets, only informal text-based descriptions of
data attributes are permitted.  In the case of EML, there is some
overlap between these two mechanisms---particularly with respect to
measurement standards---however, semantic annotations extend this
approach by providing a general mechanism to formally associate
concepts drawn from domain ontologies to attributes. In the Semtools
framework, we employ an XML serialization syntax for semantic
annotations that is compatible with EML but that is stored separately
from the EML documentation of a data set (allowing, e.g., annotations
to be used independently of EML or with other metadata standards if
needed). In addition, semantic annotations can be used to
``materialize'' a given data set into a set of triples conforming to
the model configuration given in the annotation. In other words, a
tabular data set such as the one shown in \figref{fig:kelp-mass-model}
can automatically be converted into a corresponding collection of
observation and measurement instances. This in turn enables a simple
form of structural integration, where instead of having a large number
of different tabular data structures, all data is represented using
the standard set of structures defined by the OBOE model (see
\figref{fig:oboe}). Thus, materializing a data set in this way
provides a more uniform structural representation that can make a
number of discovery and integration tasks easier. For instance,
materialization can be used to increase query expressivity by allowing
searches of the form ``\emph{find all data sets containing Mass
  measurements of Kelp with values less than or equal to 5 grams}'', which in
our example can be answered by generating (i.e., materializing) the
measurements associated with the {\tt WET} and {\tt DRY} attributes in the data set
of \figref{fig:kelp-mass-model}.

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows data owners and curators to define annotations for existing data package
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierarchical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data
packages with semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated as they fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data-set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., observations). The editor
provides a number of additional features including the ability to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements.
